---
title: "SME DataPrep R examples"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (requireNamespace("smeDataPrep", quietly = TRUE)) {
  library(smeDataPrep)
} else {
  for (file in list.files("package/R", full.names = TRUE)) {
    source(file)
  }
}
```

# Goal

This document illustrates the package with the Titanic training dataset stored in `data/titanic/train.csv`. The target variable is `Survived`, where `1` means the passenger survived and `0` means they did not.

# Data

The Titanic training file has passenger-level information and the binary outcome `Survived`. For this first example, we keep a focused set of variables so every function is easy to interpret.

```{r}
raw_data <- read.csv("data/titanic/train.csv")

analysis_data <- raw_data[c(
  "Survived",
  "Pclass",
  "Sex",
  "Age",
  "SibSp",
  "Parch",
  "Fare",
  "Embarked"
)]

analysis_data$Pclass <- as.factor(analysis_data$Pclass)
analysis_data$Survived <- as.integer(analysis_data$Survived)

dataset <- sme_dataset(analysis_data, target = "Survived")
head(analysis_data)
```

# Metrics

Numerical variables receive variance and AUC. Categorical variables receive entropy.

```{r}
metrics <- attribute_metrics(analysis_data, target = "Survived", positive_class = 1)
metrics
```

# Discretization

```{r}
numeric_features <- c("Age", "SibSp", "Parch", "Fare")

head(discretize_dataset_equal_width(analysis_data[numeric_features], bins = 4))
head(discretize_dataset_equal_frequency(analysis_data[numeric_features], bins = 4))
```

# Normalization and standardization

```{r}
head(normalize_dataset(analysis_data[numeric_features]))
head(standardize_dataset(analysis_data[numeric_features]))
```

# Filtering and associations

```{r}
filtered_by_auc <- filter_variables(
  analysis_data,
  metric = "auc",
  threshold = 0.55,
  target = "Survived",
  positive_class = 1
)
head(filtered_by_auc)

matrix <- association_matrix(analysis_data[c("Pclass", "Sex", "Age", "SibSp", "Parch", "Fare", "Embarked")])
matrix
```

# Plots

```{r}
plot_auc_values(analysis_data, target = "Survived", positive_class = 1)
plot_association_matrix(matrix)
```

# Next steps

When the second dataset is added, repeat the same workflow in a new section: load the data, identify the binary target, choose meaningful numerical/categorical variables, and show the same functions again.
