---
title: "SME DataPrep R examples"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(smeDataPrep)
```

# Example dataset

This document uses a small synthetic dataset. Later, it can be replaced with a Kaggle dataset that has numerical variables, categorical variables, and one binary target.

```{r}
data <- data.frame(
  age = c(22, 35, 58, 45, 29, 41, 63, 50),
  income = c(1200, 2500, 4200, 3900, 1800, 3100, 5100, 4700),
  segment = c("low", "medium", "high", "high", "low", "medium", "high", "high"),
  region = c("north", "north", "south", "east", "west", "east", "south", "west"),
  target = c(0, 0, 1, 1, 0, 1, 1, 1)
)

dataset <- sme_dataset(data, target = "target")
data
```

# Metrics

Numerical variables receive variance and AUC. Categorical variables receive entropy.

```{r}
attribute_metrics(data, target = "target", positive_class = 1)
```

# Discretization

```{r}
discretize_dataset_equal_width(data[c("age", "income")], bins = 3)
discretize_dataset_equal_frequency(data[c("age", "income")], bins = 3)
```

# Normalization and standardization

```{r}
normalize_dataset(data)
standardize_dataset(data)
```

# Filtering and associations

```{r}
filter_variables(data, metric = "auc", threshold = 0.7, target = "target", positive_class = 1)

matrix <- association_matrix(data[c("age", "income", "segment", "region")])
matrix
```

# Plots

```{r}
plot_auc_values(data, target = "target", positive_class = 1)
plot_association_matrix(matrix)
```
